\vspace{-2.0em} Move Schedules: \\ Fast persistence computations in sparse dynamic settings\thanks{This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.}\vspace{-0.5em}

### 2021-08-05 move_schedules.tex
Reworked overview!

\section{Introduction} 
\subsection{Overview}\label{sec:overview} 
The starting point to the work described in this paper is the stability results  established in~\cite{cohen2007stability, cohen2006vines}. 
Indeed, given a triangulable topological space equipped with a sufficiently tame continuous function, persistent homology captures the changes in topology across the sublevel sets of the space, and encodes them in a persistence diagram. The stability of persistence contends 
that small changes in the underlying function produce small changes in the corresponding diagrams:  if the function changes continuously, so too will the points on the persistence diagram. 
This motivates the study and application of persistence to time-varying or dynamic settings, like that of dynamic metric spaces~\cite{kim2020spatiotemporal}.
Focusing on the computational aspects of this problem has far-reaching consequences: methods that vectorize  persistence diagrams for machine learning purposes, like adaptive template functions~\cite{polanco2019adaptive} and persistence images~\cite{adams2017persistence}, or summary statistics like $\alpha$-smoothed Betti curves~\cite{ulmer2019topological}, all immediately become  computationally viable tools for time-varying settings. 
%This analogy provides a strong interpretation at gaining insight on the subtle topological changes that may occur in studying a continuous process.

Cohen-Steiner et al. refer to a continuous 1-parameter family of persistence diagrams as a \emph{vineyard}, and they give in \cite{cohen2006vines} an efficient algorithm for their computation
given a time-varying filtration. 
Computing vineyards can be interpreted as an extension of the \emph{matrix reduction} algorithm~\cite{zomorodian2005computing}, which computes the persistence diagram of a fixed filtration $K$ with $m$ simplices in $O(m^3)$ time.
The main step in the reduction algorithm is  a specific   decomposition $R = D V$ of the  boundary matrix $D$ of $K$. 
Since $V$ is always full rank (and upper-triangular), an alternative is to compute a $D = RU$ decomposition, where $U = V^{-1}$. 
The vineyards algorithm transforms a time-varying filtration into an ordered set of adjacent transpositions to be applied to the  rows and columns of the decomposition $D = RU$, each of which take at most $O(m)$ time to execute. If one is interested in understanding how the persistent homology of a continuous function changes over time, then this algorithm is sufficient, for  homological critical points can only occur when the filtration order changes. 
If there are $d$ time-points where this order changes, then the vineyards algorithm takes $O(m^3 + md)$ time: one $O(m^3)$-time reduction upfront for the filtration at time $t_0$ followed by one $O(m)$ operation to update the decomposition for the remaining time points $(t_1, t_2, \dots, t_d)$. When $d >> m$, this $O(md)$ approach is far more efficient than the $O(dm^3)$  naive strategy of computing the diagrams independently at every time point.

Despite the vineyards algorithm being efficient in theory, it is often not the method of choice in practice in dynamic settings. 
While there is an increasingly rich ecosystem of software packages offering variations of the standard reduction algorithm (e.g. Ripser, PHAT, Dionysus, etc... see~\cite{otter2017roadmap} for an overview), implementations of the vineyards algorithm are relatively uncommon.\footnote{Dionysus 1 has an implementation of vineyards, but the algorithm was never ported to the 2nd version.} The reason for this disparity is perhaps explained by Lesnick and Wright~\cite{lesnick2015interactive}: ``While an update to an $RU$ decomposition involving few transpositions is very fast in practice... many transpositions can be quite slow... it is sometimes much faster to simply recompute the $RU$-decomposition from scratch using the standard persistence algorithm.'' Indeed, whether performing the reductions independently or updating the decomposition dynamically using vineyards, they observed that obtaining the decomposition is the most computationally demanding aspect of their 2-D persistence algorithm.

This work seeks to further understand and remedy this discrepancy: building on the work presented in~\cite{busaryev2010tracking}, we introduce a coarser approach to the vineyards algorithm.
While the vineyards algorithm is efficient at constructing a \emph{continuous} time family of diagrams, it is not necessarily efficient when the time parameter is coarsely discretized.
Our methodology is based on the observation that practitioners often don't need (or want!) all $d$ persistence diagrams generated by a continuous 1-parameter of filtrations; usually just $n << d$ of them are sufficient.   
% in many time varying settings, a valid decomposition is often only sought after at a relatively sparse number of time points relative to the size of the (finite) set needed to generated by the family. 
By making a tradeoff in the number of times we restore the decomposition to a valid state, we are able to reduce the total number of column operations needed to apply an arbitrary permutation to the filtration. This tradeoff, paired with a fast greedy heuristic explained in section~\ref{sec:proxy_objective}, yields an algorithm that can update a $R = DV$ decomposition more efficiently than vineyards in sparse contexts, making dynamic persistence more computationally tractable for a wider class of use-cases.

@@

 \subsection{A Motivating Example and Contributions}\label{sec:motivation} 
As a motivating example, we describe a simple experiment illustrating why the vineyards algorithm does not always yield an efficient method in time-varying applications. 
Consider a series of grayscale images (i.e. a video) depicting a fixed-width annulus expanding about the center of a $9 \times 9$ grid, where scale of the annulus is parameterized by a single parameter $\tau$. Thus, the data set is a 1-parameter family of images. Each image contains the same number of pixels whose intensities---these are determined by the parameter---vary with time. For each image, we build a filtered simplicial complex using the \emph{Freudenthal} triangulation, with the ordering of inclusions obtained by lower stars.  
 \begin{figure*}[t]
 	\centering
 	\includegraphics[width=0.36\textwidth, height = 1.65in]{circle_vineyards.png}
 	\includegraphics[width=0.63\textwidth]{circle_complex.png}
 	\caption{Grayscale image data set depicting an expanding annulus (top right), each image's corresponding lower star filtration (bottom right), and the cumulative number of column operations required to compute persistence for this time-varying filtration (left). The persistent Betti numbers for each filtration from left to right are: $\beta_1 = (1,0,0), \beta_2 = (1,1,0), \beta_3 = (1,1,0), \beta_4 = (1,1,0), \beta_5 = (4,0,0)$. Observe computing $10$ diagrams independently at evenly spaced time points (the green line) both captures these major topological changes and is the most computationally efficient approach shown.}
 	\label{fig:vineyards}
 \end{figure*}
This time-varying filtration is depicted in Figure~\ref{fig:vineyards} (bottom right), which we use as a baseline for comparing the standard algorithm $\texttt{pHcol}$ with vineyards. Note the size of the filtration is fixed--no simplices are ever added or removed over time. 
There are two events which dramatically alter the persistence diagrams extracted from these filtrations: the first occurs when the central connected component splits to form a cycle, and the second occurs when the annulus approaches the outer dimensions of the grid, splitting into four components. As a consequence, in this example, only a few persistence diagrams are needed to capture the major topological changes over time.

  Consider the situation where one is concerned with capturing these dramatic changes in homology.
  On the left side of Figure~\ref{fig:vineyards}, we compare the cumulative cost (in total number of column operations) of various approaches which one can use to simulate persistence dynamically. 
  Since it is unknown a-priori at what points in  time the persistent homology will change in application settings, one solution is to discretize the time domain at $n$ evenly spaced points and compute the persistence diagram at each of these time points independently. An alternative approach is to construct a homotopy between two time-points, and then decompose this homotopy into adjacent transpositions--the vineyards approach. The former is often used in practice, but it is only an approximation; the latter is guaranteed to capture all homological changes in persistence that occur in simulating the 1-parameter family. The graph in Figure~\ref{fig:vineyards} depicts two different size approximations (purple, $n=100$ and green, $n=10$), wherein the reduction algorithm (\texttt{pHcol}) was applied at $n$ time points, and two exact strategies (black and blue), which compute all $\approx 7 \times 10^{4}$ diagrams given by the homotopy. As one can see from the figure, if one needs to simulate the homotopy exactly, the vineyards approach is indeed far more efficient than naively  applying the reduction algorithm independently at all time points.
  However,  when the discretization of the time domain is sparse enough, the naive approach   performs less column operations than the vineyards strategy, while still capturing the main events.

  The existence of a time discretization that is more efficient to compute than continually updating the decomposition indicates that the vineyards algorithm must incur some amount of overhead (in terms of column operations) to perform the transpositions. That is, even in situations where one is updating a decomposition to a filtration that is  relatively ``close'', e.g. any two adjacent filtrations in the bottom left of Figure~\ref{fig:vineyards}, it is still more efficient to apply \texttt{pHcol} directly than  to iteratively update the decomposition (see the case where $n = 10$). 
 Indeed, any optimizations to the reduction algorithm (for example, the clearing optimization~\cite{chen2011persistent}) would only increase this disparity, making the vineyards algorithm less computationally attractive even in settings where $n$ is larger. Since dynamic or time-varying are common settings where persistence is applied (see ~\cite{topaz2015topological, xian2020capturing, lesnick2015interactive, kim2020spatiotemporal}), it is worth studying this phenomenon.

\textbf{Our results and approach are as follows:} First, we leverage work of Busaryev et. al.~\cite{busaryev2010tracking} to include a coarser type of operation to the dynamic setting, called \emph{moves}. We give a tight lower bound on the number of moves needed to perform an arbitrary permutation to the $R = D V$ decomposition, and give a proof of optimality by a reduction to the permutation edit distance problem.
We also give worst-case bounds in expectation as well as efficient algorithms for achieving these bounds---both of which are derived from a reduction to the Longest Increasing Subsequence (LIS) problem. 
This reduction yields an efficient algorithm for generating sequences of moves $( s_1, s_2, \dots, s_d )$, which we call \emph{schedules}. 
We investigate the feasibility of optimizing the cost 
(i.e., number of column operations) of these schedules directly, including the existence of a greedy-type approach, which we show can lead to arbitrarily bad behavior. 
In light of these results, we give an alternative proxy-objective to minimize, provide bounds justifying its relevance to the original objective, and give an efficient $O(d \log d)$ algorithm for approximately solving this proxy minimization. 
A performance comparison with other reduction-based persistent homology computations is given, wherein our approach is demonstrated to be an order of magnitude more efficient than existing approaches in sparse dynamic settings. We conclude with a discussion of situations where our strategy is most applicable and future directions for research.

@@

In contrast, there is extensive work  on improving the efficiency of computing a (static) $R = DV$ decomposition. Chen~\cite{chen2011persistent} proposed \emph{persistence with a twist}, also called the \emph{clearing optimization}, which exploits a boundary/cycle relationship to ``kill'' columns early in the reduction rather than reducing them. 
Another popular optimization is to utilize the duality between   homology and cohomology~\cite{de2011dualities}, which dramatically improves the effectiveness of the clearing optimization~\cite{bauer2021ripser}. 
There are many other optimizations on the implementation side: the use of ranking functions defined on the combinatorial number system enables implicit cofacet enumeration, removing the need to store the boundary matrix explciitly; the apparent/emergent pairs optimization identifies columns whose pivot entries are unaffected by the reduction algorithm, reducing the total number of columns which need be reduced; sparse data structures such as bit-trees and lazy heaps allow for efficient column-wise additions with $\mathbb{Z}_2 = \mathbb{Z}/2\mathbb{Z}$ coefficients and effective $O(1)$ pivot entry retrieval, and so on~\cite{bauer2017phat, bauer2021ripser}.

@@

\subsection{Outline} The remainder of the paper is organized as follows: we review and establish the notation we will use to describe simplicial complexes, persistent homology, and dynamic persistence. We also cover the reduction algorithm (designated here as \texttt{pHcol}), the vineyards algorithm, and the set of \emph{move}-related algorithms introduced in~\cite{busaryev2010tracking}, which serves as the starting point of this work. 
In section 3 we introduce our proposed alternative to vineyards and provide efficient algorithms to make this alternative viable. 
In section 4 we demonstrate a few applications of the proposed method. In section 5 we conclude the paper by discussing other possible applications, future work. 
We also include an appendix which introduces certain topics in more detail, such as the vineyards algorithm itself, for readers who may be unfamiliar but would like a more in-depth understanding.

@@

\begin{corollary}\label{cor:expectation}
	If $K_0, K_1$ are random distinct filtrations each of size $m$, then $d$ is upper bounded by $O(m - \sqrt{m})$ in expectation, with probability $1$ as $m \to \infty$.
\end{corollary}
\begin{proof}
	The proof of this result reduces to showing the average length of the LIS for random permutations. Let $L(p) \in [1,m]$ denote the maximal length of a increasing subsequence of $p \in S_m$. 
	The essential quantity to show the expected length of $L(p)$ over all permutations: 
	$$ \ell_m = \mathbb{E} \, L(p) = \frac{1}{m!} \sum\limits_{p \in S_m} L(p)$$
	The history of the mathematics of estimating this quantity dates back at least 50 years, and is sometimes called the \emph{Ulam-Hammersley} problem. It was shown by Baik et al.~\cite{} that as $m \to \infty$:
	$$ \displaystyle \ell_m = 2 \sqrt{m} + c m^{1/6} + o(m^{1/6}) $$
where $c \approx -1.77108$. As a result, we have: 
$$ \frac{\ell_m}{\sqrt{m}} \to 2\rlap{\quad \text{as } $m \to \infty$} $$	 
Thus, if $p \in S_m$ denotes a uniformly random permutation in $S_m$, then $L(p)/\sqrt{m} \to 2$ in probability as $m \to \infty$. 
Since the schedule size is defined as $d = m - \lvert \mathrm{LCS}(K_0, K_1) \rvert$, the claimed bound follows.
\end{proof}
\noindent The upper bound from Corollary~\ref{cor:expectation} captures the size of $S^\ast$ under worst-case conditions. It is worth noting that this bound applies to pairs of uniformly sampled permutations, as opposed to uniformly sampled distinct filtrations---intuitively, simplicial filtrations have more structure due to the subcomplex relation that must be respected in the filtration. Thus, while $O(m - \sqrt{m})$ in expectation is an improvement over $O(m)$ it is still not tight. However, Boissonnat~\cite{boissonnat2018efficient} has shown that for any $d$-dimensional simplicial complex $K$ with $m$ simplices, the number of distinct filtrations built from $K$ is \emph{at least} $\floor{\frac{t+1}{d+1}}^m$, where $t$ is the number of distinct filtration values associated with the complex $K$. Since this lower bound grows similarly to $m!$ when $t \approx O(m)$ and $d << m$ fixed, we expect $O(m - \sqrt{m})$ is not too pessimistic of an upper bound when the pair of filtrations are truly random. In practice, when one has a time-varying filtration and the sampling points are relatively close [in time], the LCS between adjacent filtrations is expected to be much larger, subsequently making $d$ much smaller. 
This had been confirmed empirically, e.g. the average size of the LCS across the 10 filtrations from Section~\ref{sec:motivation} each having $417$ simplices sampled at evenly spaced time points was $343$, implying $d \approx 70$ permutations needed on average to update the decomposition between adjacent time points.

>>>>

\section{Introduction} 
\subsection{Overview}\label{sec:overview} 
The starting point to the work described in this paper is the stability results established in~\cite{cohen2007stability, cohen2006vines}. 
Indeed, given a triangulable topological space equipped with a sufficiently tame continuous function, persistent homology captures the changes in topology across the sublevel sets of the space, and encodes them in a persistence diagram. The stability of persistence contends that small changes in the underlying function produce small changes in the corresponding diagrams:  if the function changes continuously, so too will the points on the persistence diagram. 
This motivates the study and application of persistence to time-varying or dynamic settings, like that of dynamic metric spaces~\cite{kim2020spatiotemporal}. 
As persistence-related computations tend to exhibit high algorithmic complexity (e.g. essentially cubic\footnote{For finite fields, it is known that the persistence computation reduces to the PLU factorization problem, which takes $O(m^\omega)$ where $\omega \approx 2.373$ is the matrix multiplication constant.} in the size of the underlying filtration~\cite{morozov2005persistence}), their adoption to dynamic settings poses a challenging computational problem.
With the current state of the art tools, there is no recourse to the problem of computing persistence from a time-varying filtration containing millions of simplices at thousands of snapshots in time.
An efficient means of updating the persistence of a [time-varying] filtration has far-reaching consequences: methods that vectorize persistence diagrams for machine learning purposes, like adaptive template functions~\cite{polanco2019adaptive} and persistence images~\cite{adams2017persistence}, or summary statistics like $\alpha$-smoothed Betti curves~\cite{ulmer2019topological}, all immediately become computationally viable tools in dynamic  settings. 
%This analogy provides a strong interpretation at gaining insight on the subtle topological changes that may occur in studying a continuous process.

Cohen-Steiner et al. refer to a continuous 1-parameter family of persistence diagrams as a \emph{vineyard}, and they give in \cite{cohen2006vines} an efficient algorithm for their computation given a time-varying filtration. 
The vineyards approach can be interpreted as an extension of the \emph{matrix reduction} algorithm~\cite{zomorodian2005computing}, which computes the persistence diagram of a fixed filtration $K$ with $m$ simplices in $O(m^3)$ time.
The main step in the reduction algorithm is a specific decomposition $R = D V$ of the boundary matrix $D$ of $K$. 
Since $V$ is always full rank (and upper-triangular), an alternative is to compute a $D = RU$ decomposition, where $U = V^{-1}$. 
The vineyards algorithm transforms a time-varying filtration into an ordered set of adjacent transpositions to be applied to the rows and columns of the decomposition $D = RU$, each of which take at most $O(m)$ time to execute. If one is interested in understanding how the persistent homology of a continuous function changes over time, then this algorithm is sufficient, for  homological critical points can only occur when the filtration order changes. 
Indeed, this algorithm is also efficient  asymptotically: if there are $d$ time-points where this filtration order changes, then the vineyards algorithm takes $O(m^3 + md)$ time: one $O(m^3)$-time reduction upfront for the filtration at time $t_0$ followed by one $O(m)$ operation to update the decomposition for the remaining time points $(t_1, t_2, \dots, t_d)$. When $d >> m$, this $O(md)$ approach is far more efficient than the $O(dm^3)$  strategy of computing the diagrams independently at every time point, what we call the ``naive'' approach.

Despite its efficient in theory, vineyards is often not the method of choice in practice in dynamic settings. 
While there is an increasingly rich ecosystem of software packages offering variations of the standard reduction algorithm (e.g. Ripser, PHAT, Dionysus, etc... see~\cite{otter2017roadmap} for an overview), implementations of the vineyards algorithm are relatively uncommon.\footnote{Dionysus 1 has an implementation of vineyards, but the algorithm was never ported to the 2nd version.} The reason for this disparity is perhaps explained by Lesnick and Wright~\cite{lesnick2015interactive}: ``While an update to an $RU$ decomposition involving few transpositions is very fast in practice... many transpositions can be quite slow... it is sometimes much faster to simply recompute the $RU$-decomposition from scratch using the standard persistence algorithm.'' Indeed, whether performing the reductions naively or updating the decomposition dynamically using vineyards, they observed that obtaining the decomposition is the most computationally demanding aspect of their 2-D persistence algorithm.

This work seeks to further understand and remedy this discrepancy: building on the work presented in~\cite{busaryev2010tracking}, we introduce a coarser approach to the vineyards algorithm.
While the vineyards algorithm is efficient at constructing a \emph{continuous} time family of diagrams, it is not necessarily efficient when the time parameter is coarsely discretized.
Our methodology is based on the observation that practitioners often don't need (or want!) all $d$ persistence diagrams generated by a continuous 1-parameter of filtrations; usually just $n << d$ of them are sufficient.   
% in many time varying settings, a valid decomposition is often only sought after at a relatively sparse number of time points relative to the size of the (finite) set needed to generated by the family. 
By exploiting the ``donor'' concept introduced in in~\cite{busaryev2010tracking}, we are able to make a tradeoff in the number of times the decomposition is restored to a valid state, dramatically reducing the total number of column operations needed to apply an arbitrary permutation to the filtration. This tradeoff, paired with a fast greedy heuristic explained in section~\ref{sec:proxy_objective}, yields an algorithm that can update a $R = DV$ decomposition more efficiently than vineyards in sparse contexts, making dynamic persistence more computationally tractable for a wider class of use-cases.

@@

 \subsection{A Motivating Example and Contributions}\label{sec:motivation} 
As a motivating example, we describe a simple experiment illustrating why the vineyards algorithm does not always yield an efficient method in many time-varying applications. 
Consider a series of grayscale images (i.e. a video) depicting a fixed-width annulus expanding about the center of a $9 \times 9$ grid, where scale of the annulus is parameterized by a single parameter $\tau$. Thus, the data set is a 1-parameter family of images. Each image contains the same number of pixels whose intensities---these are determined by the parameter---vary with time. For each image, we build a filtered simplicial complex using the \emph{Freudenthal} triangulation, with the ordering of inclusions obtained by lower stars.  
 \begin{figure*}[t]
 	\centering
 	\includegraphics[width=0.36\textwidth, height = 1.65in]{circle_vineyards.png}
 	\includegraphics[width=0.63\textwidth]{circle_complex.png}
 	\caption{Grayscale image data set depicting an expanding annulus (top right), each image's corresponding lower star filtration (bottom right), and the cumulative number of column operations required to compute persistence for this time-varying filtration (left). The persistent Betti numbers for each filtration from left to right are: $\beta_1 = (1,0,0), \beta_2 = (1,1,0), \beta_3 = (1,1,0), \beta_4 = (1,1,0), \beta_5 = (4,0,0)$. Observe computing $10$ diagrams independently at evenly spaced time points (the green line) both captures these major topological changes and is the most computationally efficient approach shown.}
 	\label{fig:vineyards}
 \end{figure*}
Snapshots of the image data and its corresponding time-varying filtration is depicted in Figure~\ref{fig:vineyards} (top right and bottom right, respectively), which we use as a baseline for comparing the standard algorithm $\texttt{pHcol}$ with vineyards. Note the size of the filtration is fixed--no simplices are ever added or removed over time. 
There are two events which dramatically alter the persistence diagrams extracted from these filtrations: the first occurs when the central connected component splits to form a cycle, and the second occurs when the annulus approaches the outer dimensions of the grid, splitting into four components. As a consequence, in this example, only a few persistence diagrams are needed to capture the major topological changes over time.

  Consider the situation where one is concerned with capturing these dramatic changes in homology.
  On the left side of Figure~\ref{fig:vineyards}, we compare the cumulative cost (in total number of column operations) of various approaches which one can use to simulate persistence dynamically. 
  Since it is unknown a priori at what points in time the persistent homology will change in application settings, one solution is to discretize the time domain at $n$ evenly spaced points and compute the persistence diagram at each of these time points independently. An alternative approach is to construct a homotopy between two time-points, and then decompose this homotopy into adjacent transpositions--the vineyards approach. The former is often used in practice, but it is only an approximation; the latter is guaranteed to capture all homological changes in persistence that occur in simulating the 1-parameter family. The graph in Figure~\ref{fig:vineyards} depicts two different size approximations (purple, $n=100$ and green, $n=10$), wherein the reduction algorithm (\texttt{pHcol}) was applied at $n$ time points, and two exact strategies (black and blue), which compute all $\approx 7 \times 10^{4}$ diagrams given by the homotopy. As one can see from the figure, if one needs to simulate the homotopy exactly, the vineyards approach is indeed far more efficient than naively applying the reduction algorithm independently at all time points.
  However, when the discretization of the time domain is sparse enough, the naive approach actually performs less column operations than the vineyards strategy, while still capturing the main events.

  The existence of a time discretization that is more efficient to compute than continually updating the decomposition indicates that the vineyards algorithm must incur some amount of overhead (in terms of column operations) to perform the transpositions. That is, even in situations where one is updating a decomposition to a filtration that is  relatively ``close'', e.g. any two adjacent filtrations in the bottom right of Figure~\ref{fig:vineyards}, it is still more efficient to apply \texttt{pHcol} directly than  to iteratively update the decomposition (see the case where $n = 10$). 
 Indeed, any optimizations to the reduction algorithm (for example, the clearing optimization~\cite{chen2011persistent}) would only increase this disparity, making the vineyards algorithm less computationally attractive even in settings where $n$ is larger. Since dynamic or time-varying are common settings where persistence is applied (see ~\cite{topaz2015topological, xian2020capturing, lesnick2015interactive, kim2020spatiotemporal}), it is worth studying this phenomenon.

\textbf{Our results and approach are as follows:} First, we leverage work of Busaryev et. al.~\cite{busaryev2010tracking} to include a coarser type of operation to the dynamic setting, called \emph{moves}. We give a tight lower bound on the number of moves needed to perform an arbitrary permutation to the $R = D V$ decomposition, and give a proof of optimality by a reduction to the permutation edit distance problem.
We also give worst-case bounds in expectation as well as efficient algorithms for achieving these bounds---both of which are derived from a reduction to the Longest Increasing Subsequence (LIS) problem. 
This reduction yields an efficient algorithm for generating sequences of moves $( s_1, s_2, \dots, s_d )$, which we call \emph{schedules}. 
We investigate the feasibility of optimizing the cost 
(i.e., number of column operations) of these schedules directly, including the existence of a greedy-type approach, which we show can lead to arbitrarily bad behavior. 
In light of these results, we give an alternative proxy-objective to minimize, provide bounds justifying its relevance to the original objective, and give an efficient $O(d \log d)$ algorithm for approximately solving this proxy minimization. 
A performance comparison with other reduction-based persistent homology computations is given, wherein our approach is demonstrated to be an order of magnitude more efficient than existing approaches in a few common sparse dynamic settings, such as flock simulation and 2D persistence computation. We conclude with a discussion of situations where our strategy is most applicable and future directions for research.

@@

In contrast, there is extensive work  on improving the efficiency of computing a (static) $R = DV$ decomposition. Chen~\cite{chen2011persistent} proposed \emph{persistence with a twist}, also called the \emph{clearing optimization}, which exploits a boundary/cycle relationship to ``kill'' columns early in the reduction rather than reducing them. 
Another popular optimization is to utilize the duality between   homology and cohomology~\cite{de2011dualities}, which dramatically improves the effectiveness of the clearing optimization~\cite{bauer2021ripser}. 
There are many other optimizations on the implementation side: the use of ranking functions defined on the combinatorial number system enables implicit cofacet enumeration, removing the need to store the boundary matrix explciitly; the apparent/emergent pairs optimization identifies columns whose pivot entries are unaffected by the reduction algorithm, reducing the total number of columns which need be reduced; sparse data structures such as bit-trees and lazy heaps allow for efficient column-wise additions with $\mathbb{Z}_2 = \mathbb{Z}/2\mathbb{Z}$ coefficients and effective $O(1)$ pivot entry retrieval, and so on~\cite{bauer2021ripser, bauer2017phat}.

@@

\subsection{Outline} The remainder of the paper is organized as follows: we review and establish the notation we will use to describe simplicial complexes, persistent homology, and dynamic persistence. We also cover the reduction algorithm (designated here as \texttt{pHcol}), the vineyards algorithm, and the set of \emph{move}-related algorithms introduced in~\cite{busaryev2010tracking}, which serves as the starting point of this work. 
In section 3 we introduce our proposed alternative to vineyards and provide efficient algorithms to make this alternative viable. 
In section 4 we demonstrate a few applications of the proposed method. In section 5 we conclude the paper by discussing other possible applications and future work. 
We also include an appendix which introduces certain topics in more detail, such as the vineyards algorithm itself, for readers who may be unfamiliar but would like a more in-depth understanding.

@@

\begin{corollary}\label{cor:expectation}
	If $K_0, K_1$ are random distinct filtrations each of size $m$, then $d$ is upper bounded by $O(m - \sqrt{m})$ in expectation, with probability $1$ as $m \to \infty$.
\end{corollary}
\begin{proof}
	The proof of this result reduces to showing the average length of the LIS for random permutations. Let $L(p) \in [1,m]$ denote the maximal length of a increasing subsequence of $p \in S_m$. 
	The essential quantity to show the expected length of $L(p)$ over all permutations: 
	$$ \ell_m = \mathbb{E} \, L(p) = \frac{1}{m!} \sum\limits_{p \in S_m} L(p)$$
	The history of the mathematics of estimating this quantity dates back at least 50 years, and is sometimes called the \emph{Ulam-Hammersley} problem. It was shown by Baik et al.~\cite{} that as $m \to \infty$:
	$$ \displaystyle \ell_m = 2 \sqrt{m} + c m^{1/6} + o(m^{1/6}) $$
where $c \approx -1.77108$. As a result, we have: 
$$ \frac{\ell_m}{\sqrt{m}} \to 2\rlap{\quad \text{as } $m \to \infty$} $$	 
Thus, if $p \in S_m$ denotes a uniformly random permutation in $S_m$, then $L(p)/\sqrt{m} \to 2$ in probability as $m \to \infty$. 
Since the schedule size is defined as $d = m - \lvert \mathrm{LCS}(K_0, K_1) \rvert$, the claimed bound follows.
\end{proof}
\noindent The upper bound from Corollary~\ref{cor:expectation} captures the size of $S^\ast$ under worst-case conditions. It is worth noting that this bound applies to pairs of uniformly sampled permutations, as opposed to uniformly sampled distinct filtrations---intuitively, simplicial filtrations have more structure due to the subcomplex relation that must be respected in the filtration. Thus, while $O(m - \sqrt{m})$ in expectation is an improvement over $O(m)$, it is still not tight. However, Boissonnat~\cite{boissonnat2018efficient} has shown that for any $d$-dimensional simplicial complex $K$ with $m$ simplices, the number of distinct filtrations built from $K$ is \emph{at least} $\floor{\frac{t+1}{d+1}}^m$, where $t$ is the number of distinct filtration values associated with the complex $K$. Since this lower bound grows similarly to $m!$ when $t \approx O(m)$ and $d << m$ fixed, we expect $O(m - \sqrt{m})$ is not too pessimistic of an upper bound when the pair of filtrations are truly random. In practice, when one has a time-varying filtration and the sampling points are relatively close [in time], the LCS between adjacent filtrations is expected to be much larger, subsequently making $d$ much smaller. 
For example, the average size of the LCS across the 10 filtrations from Section~\ref{sec:motivation} each having $m=417$ simplices sampled at evenly spaced time points was $343$, implying $d \approx 70$ permutations needed on average to update the decomposition between adjacent time points.
